Architectures and Open pNets

### 2018-10-20 2018-Arch-pNets/arch-pNets.tex
Arch-pNets paper: Finished a first draft of the encoding without priorities

\usepackage{amsmath,amsthm,amscd}
\usepackage{amssymb}    % for blackboard B
\usepackage{rotating}   % to rotate the figures
\usepackage{mathpartir} % math paragraph + inference rules
\usepackage{mathtools}  % for align, multline etc.
\usepackage{caption}
\usepackage{stmaryrd} % double brackets (\llbracket, \rrbracket)

@@

@@

\begin{definition}[pNets]\label{def-pnets}
A pNet is a hierarchical structure where leaves are pLTSs and holes\footnote{Typing can 
be added to pNets by defining and checking their interfaces, called 
sort~\cite{HMZ-FORTE2016}}:
$\pNet\triangleq \pLTS~|~\mylangle \pNet_i^{i\in I}, J, \symb{SV}_k^{k\in 
K}\myrangle$
where
\begin{itemize}
\item[$\bullet$] $\pNet_i^{i\in I}$ is the family of sub-pNets;
%  $\pNet_i^{i\in I}$ is a family of sub-pNets where $I\in\I_\P$ is the set over which 
%sub-pNets are indexed.

@@

\noteLH{a pred is a g a post is a e}
The semantics of open pNets will be defined  as an open automaton. An open
automaton is an automaton where each transition composes the actions of several holes, 
the transition occurs if some predicates hold, and can involve a 
set of state modifications.
%\TODO{adopt a uniform notation for open transitions, almost each instance has a 
%different 
%notation! I suggest p,l,pr,po using \{\} for p,l,po as they are sets}
\begin{definition}[Open transitions]
	\label{def:OpenTransitions}
	An \emph{open transition} over a
	set $J$ of holes  and a set of states $\mathcal{S}$ is 
	a structure of the form:	
	\begin{mathpar}
	\inferrule*[myfraction=\reddottedrule]
	{	\beta_j^{j\in J}, \Pred, \Post}
	{s \OTarrow {{\alpha}}s'}
	\end{mathpar}
	Where $s, s'\in\mathcal{S}$ and $\beta_j\in\actions{\variables}$
        is an action of the hole $j$. $v$ is a variable denoting the resulting action
        of this open transition. \Pred\ is a predicate 
	over the different variables of the
	terms, labels, and states  $\beta_j$, $s$, $\alpha$. $\Post\in 
	\assigns{\variables}$\ is 
	a set of assignments that are the effects of the transition.
Open transitions are identified modulo logical equivalence on their predicate.
\end{definition}
%\begin{definition}[Open transitions]
%	\label{def:OpenTransitions}
%	An \emph{open transition} over a set \noteSB{In the architecture section, I say that 
%we omit indices on $\goesto{}$ and that they are always clear from the 
%context.}\hl{$(S_i,s_{0 i}, \rightarrow_i)^{i\in
%	I}$} of LTSs, a
%	set $J$ of holes with sorts $\Sort_j^{j\in J}$, and a set of states $\mathcal{S}$ is 
%	a structure of the form:	
%	\begin{mathpar}
%	\inferrule*[myfraction=\reddottedrule]
%	{\{s_i~{\xrightarrow{a_i}}_i ~s_i^{\prime}\}^{i\in I},
%		\{\xrightarrow{b_j}_j\}^{j\in J}, \Pred, \Post}
%	{s \OTarrow {v}s'}
%	\end{mathpar}
%	Where $s, s'\in\mathcal{S}$ and for all
%        $i\in I$, $s_i{\xrightarrow{a_i}}_i s_i^{\prime}$ is a transition of the
%	LTS $(S_i,s_{0 i}, \rightarrow_i)$, and \noteSB{Why insist on ``transition'' and 
%$\goesto{b_j}_j$? Why not simply ``$b_j$ is an action in the sort 
%$\Sort_j$?}\hl{$\xrightarrow{b_j}_j$
%        is a transition of the hole $j$}, for any action $b_j$ in the
%        sort $\Sort_j$.  
%        of this open transition. \Pred\ is a predicate 
%	over the different variables of the
%	terms, labels, and states $s_i$, $b_j$, $s$, $v$. \Post\ is a set of equations that 
%	hold \emph{after the open transition}, they are represented as a substitution of the 
%	form $\{x_k\gets \Expr_k\}^{k\in K}$ \noteLH{considering the level of abstraction I 
%	would like to just put $\Post_=e$}
%	where $x_k$ are variables of $s'$, $s'_i$, and $e_k$ are expressions over the other 
%	variables of the open transition.
%\end{definition}

@@

\begin{definition}[Open automaton]
	\label{def:open-automaton}
	An \emph{open automaton} is a structure\\ $A =
	<J,\mathcal{S},s_0,\mathcal{T}>$ where:
	\begin{itemize}
		\item[$\bullet$]   $J$ is a  set of indices,
		\item[$\bullet$]   $\mathcal{S}$ is a set of states and $s_0$ an initial state
		among $\mathcal{S}$,
		\item[$\bullet$] $\mathcal{T}$ is a set of open transitions and for each
		$t\in \mathcal{T}$ there exist  $J'$ with  $J'
		\subseteq J$, such that $t$ is an open transition over  $J'$,
		and  $\mathcal{S}$.

@@

\begin{definition}[Operational semantics of open pNets]
	\label{def:operationalSemantics}
	The semantics of a pNet $p$ is an open automaton $A = 
	<Holes(p),States(p),InitState(p),
	\mathcal{T}>$ where $\mathcal{T}$ is the smallest set of open transitions		
	satisfying the rules below:
%	\begin{itemize}
%		\item $J$ is the set of holes: $Holes(p)= J$. 
		%  \item $\overline{L}^L = Leaves(p), \overline{H}^J = Holes(p)$
%		\item ${\mathcal{S}} = States(p)$ and $s_0 = InitState(p)$
%		\item $\mathcal{T}$ is the smallest set of open transitions		satisfying the 
%rules below:
%	\end{itemize}

@@

The general case encoding relies on the causal semantics of the
algebra of BIP connectors \cite{BliSif10-causal-fmsd}.  Disregarding
the variables and data transfer, the Algebra of Connectors $\ac(P)$
\cite{BliSif07-acp-emsoft} provides a syntactic notation for the BIP
connectors.  The causal semantics of the connectors, given in terms of
the Algebra of Causal Interaction Trees $\ct(P)$, explicitates the
causal dependencies discussed above, through an encoding mapping
$\tau: \ac(P) \rightarrow \ct(P)$.  Another mapping $R: \ct(P)
\rightarrow \cru(P)$ encodes causal interaction trees into systems of
causal rules, which are Boolean implications similar to the ones in
the example above.  We briefly summarise the key definitions and
results in \app{algebras}.

@@

First, consider a flat connector with $n$ trigger ports $p_1,\dots,
p_n$ (with $n > 0$) and $m$ synchron ports $q_1,\dots, q_m$
(\fig{connector}).  Similarly to the example above, for any of the
ports $q_1,\dots, q_m$ to participate in an interaction defined by
this connector, at least one of the ports $p_1,\dots, p_n$ must
participate.  This causal dependency can be represented by the Boolean
formula $q_1 \lor \dots \lor q_m \Rightarrow p_1 \lor \dots \lor p_n$.
If there are no synchrons, \ie $m = 0$, this formula simplifies to
$p_1 \lor \dots \lor p_n$.  However, if there are no triggers, \ie $n
= 0$, the connector semantics is that \emph{all} synchron ports must
participate for an interaction to take place, represented by the
Boolean formula $(q_1 = q_2) \land \dots \land (q_1 = q_m)$.

We now generalise this principle to hierarchical connectors, building
the formula bottom-up.  In a flat connector above, we have implicitly
associated a Boolean variable to each port.  In a hierarchical
connector, we associate a fresh Boolean variable to each node.
Similarly to the above example, consider a connector with $n > 0$
trigger sub-connectors and $m$ synchron sub-connectors.  We associate
to this connector a fresh Boolean variable $x$.  Assume that Boolean
variables associated to trigger sub-connectors are $x_1,\dots,x_n$,
whereas those associated to synchron sub-connectors are
$y_1,\dots,y_m$.  The causal dependencies among the interactions
defined by the sub-connectors can be encoded by the formula
%
\begin{equation}
  \label{eq:trig:encoding}
  (y_1 \lor \dots \lor y_m \Rightarrow x) \land (x = x_1 \lor \dots
  \lor x_n).
\end{equation}
%
If there are no trigger sub-connectors, \ie $n = 0$, we use the
formula
%
\begin{equation}
  \label{eq:synch:encoding}
  x = \bigl((y_1 = y_2) \land \dots \land (y_1 = y_m)\bigr)
\end{equation}
%
($x$ is equal to $\true$ if all $x_1, \dots, x_m$ are equal, it is
equal to $\false$ otherwise).

Notice that, in the flat connector example above, the formula was
essentially the same, with $p_i$ instead of $x_i$ and $q_i$ instead of
$y_i$, and without the additional variable $x$.  Indeed, $x$ is used
to encode the causal dependencies between this connector and, if
necessary, those in higher levels of the hierarchy.  The Boolean
encoding of the causal dependencies among the ports of the entire
connector is obtained by taking the conjunction of formulas obtained
for each sub-connector.

Consider again the connector {\NameTimer.\PortTick \trigsynch
  (\PortFail \trigsynch \NameCtrl.\PortFail)}.  Associating the
varaibles $b_0$, $b_1$ and $b_2$ to ports {\PortFail},
{\NameCtrl.\PortFail} and {\NameTimer.\PortTick}, respectively, and
variables $y_1$ and $y_2$ to the top- ({\NameTimer.\PortTick
  \trigsynch $\cdot$}) and second-level (\PortFail \trigsynch \NameCtrl.\PortFail) connectors in the hierarchy,
we obtain the following formula
%
\begin{equation}
  \label{eq:tff:example}
  (b_1 \Rightarrow y_2) \land (y_2 = b_0) \land (y_2 \Rightarrow y_1)
  \land (y_1 = b_2)
  \,.
\end{equation}
%
Observe now that the top-level connector, by its very nature, does not
have any higher level connectors and therefore we can eliminate the
corresponding Boolean variable.  This simplifies \eq{tff:example}, to
%
\begin{equation}
  \label{eq:tff:example:1}
  (b_1 \Rightarrow y_2) \land (y_2 = b_0) \land (y_2 \Rightarrow b_2)
  \,.
\end{equation}
%
In the general case, this corresponds to replacing, \emph{for
  top-level connectors}, equations \eq{trig:encoding} and
\eq{synch:encoding}, respectively, by
%
\begin{gather}
  \label{eq:trig:encoding:1}
  y_1 \lor \dots \lor y_m \Rightarrow x_1 \lor \dots
  \lor x_n
  \\
  \intertext{and}
  \label{eq:synch:encoding:1}
  (y_1 = y_2) \land \dots \land (y_1 = y_m)
  \,.
\end{gather}
%
Furthermore, in the context of one connector (or synchronisation
vector), some interaction must be fired.  Thus, \emph{for top-level
  connectors}, \eq{synch:encoding:1} can be further simplified to 
%
\begin{equation}
  \label{eq:synch:encoding:2}
  y_1 \land \dots \land y_m
  \,.
\end{equation}

>>>>

\usepackage{amsmath,amsthm,amscd}
\usepackage{amssymb}    % for blackboard B
\usepackage{rotating}   % to rotate the figures
\usepackage{mathpartir} % math paragraph + inference rules
\usepackage{mathtools}  % for align, multline etc.
\usepackage{caption}
\usepackage{stmaryrd} % double brackets (\llbracket, \rrbracket)
\usepackage{units}    % for nicefrac: x/y

@@

\newcommand{\Tuple}[2][\ensuremath{,\allowbreak}]{%
  \setTupleDelims[\Big(]{\Big)}%
  \tuple[#1]{#2}%
}

@@

\begin{definition}[pNets]\label{defn:pnets}
A pNet is a hierarchical structure where leaves are pLTSs and holes\footnote{Typing can 
be added to pNets by defining and checking their interfaces, called 
sort~\cite{HMZ-FORTE2016}}:
$\pNet\triangleq \pLTS~|~\mylangle \pNet_i^{i\in I}, J, \symb{SV}_k^{k\in 
K}\myrangle$
where
\begin{itemize}
\item[$\bullet$] $\pNet_i^{i\in I}$ is the family of sub-pNets;
%  $\pNet_i^{i\in I}$ is a family of sub-pNets where $I\in\I_\P$ is the set over which 
%sub-pNets are indexed.

@@

\noteLH{a pred is a g a post is a e}
The semantics of open pNets will be defined  as an open automaton. An open
automaton is an automaton where each transition composes the actions of several holes, 
the transition occurs if some predicates hold, and can involve a 
set of state modifications.
%\TODO{adopt a uniform notation for open transitions, almost each instance has a 
%different 
%notation! I suggest p,l,pr,po using \{\} for p,l,po as they are sets}
\begin{definition}[Open transitions]
	\label{defn:OpenTransitions}
	An \emph{open transition} over a
	set $J$ of holes  and a set of states $\mathcal{S}$ is 
	a structure of the form:	
	\begin{mathpar}
	\inferrule*[myfraction=\reddottedrule]
	{	\beta_j^{j\in J}, \Pred, \Post}
	{s \OTarrow {{\alpha}}s'}
	\end{mathpar}
	Where $s, s'\in\mathcal{S}$ and $\beta_j\in\actions{\variables}$
        is an action of the hole $j$. $v$ is a variable denoting the resulting action
        of this open transition. \Pred\ is a predicate 
	over the different variables of the
	terms, labels, and states  $\beta_j$, $s$, $\alpha$. $\Post\in 
	\assigns{\variables}$\ is 
	a set of assignments that are the effects of the transition.
Open transitions are identified modulo logical equivalence on their predicate.
\end{definition}
%\begin{definition}[Open transitions]
%	\label{def:OpenTransitions}
%	An \emph{open transition} over a set \noteSB{In the architecture section, I say that 
%we omit indices on $\goesto{}$ and that they are always clear from the 
%context.}\hl{$(S_i,s_{0 i}, \rightarrow_i)^{i\in
%	I}$} of LTSs, a
%	set $J$ of holes with sorts $\Sort_j^{j\in J}$, and a set of states $\mathcal{S}$ is 
%	a structure of the form:	
%	\begin{mathpar}
%	\inferrule*[myfraction=\reddottedrule]
%	{\{s_i~{\xrightarrow{a_i}}_i ~s_i^{\prime}\}^{i\in I},
%		\{\xrightarrow{b_j}_j\}^{j\in J}, \Pred, \Post}
%	{s \OTarrow {v}s'}
%	\end{mathpar}
%	Where $s, s'\in\mathcal{S}$ and for all
%        $i\in I$, $s_i{\xrightarrow{a_i}}_i s_i^{\prime}$ is a transition of the
%	LTS $(S_i,s_{0 i}, \rightarrow_i)$, and \noteSB{Why insist on ``transition'' and 
%$\goesto{b_j}_j$? Why not simply ``$b_j$ is an action in the sort 
%$\Sort_j$?}\hl{$\xrightarrow{b_j}_j$
%        is a transition of the hole $j$}, for any action $b_j$ in the
%        sort $\Sort_j$.  
%        of this open transition. \Pred\ is a predicate 
%	over the different variables of the
%	terms, labels, and states $s_i$, $b_j$, $s$, $v$. \Post\ is a set of equations that 
%	hold \emph{after the open transition}, they are represented as a substitution of the 
%	form $\{x_k\gets \Expr_k\}^{k\in K}$ \noteLH{considering the level of abstraction I 
%	would like to just put $\Post_=e$}
%	where $x_k$ are variables of $s'$, $s'_i$, and $e_k$ are expressions over the other 
%	variables of the open transition.
%\end{definition}

@@

\begin{definition}[Open automaton]
	\label{defn:open-automaton}
	An \emph{open automaton} is a structure\\ $A =
	<J,\mathcal{S},s_0,\mathcal{T}>$ where:
	\begin{itemize}
		\item[$\bullet$]   $J$ is a  set of indices,
		\item[$\bullet$]   $\mathcal{S}$ is a set of states and $s_0$ an initial state
		among $\mathcal{S}$,
		\item[$\bullet$] $\mathcal{T}$ is a set of open transitions and for each
		$t\in \mathcal{T}$ there exist  $J'$ with  $J'
		\subseteq J$, such that $t$ is an open transition over  $J'$,
		and  $\mathcal{S}$.

@@

\begin{definition}[Operational semantics of open pNets]
	\label{defn:operationalSemantics}
	The semantics of a pNet $p$ is an open automaton $A = 
	<Holes(p),States(p),InitState(p),
	\mathcal{T}>$ where $\mathcal{T}$ is the smallest set of open transitions		
	satisfying the rules below:
%	\begin{itemize}
%		\item $J$ is the set of holes: $Holes(p)= J$. 
		%  \item $\overline{L}^L = Leaves(p), \overline{H}^J = Holes(p)$
%		\item ${\mathcal{S}} = States(p)$ and $s_0 = InitState(p)$
%		\item $\mathcal{T}$ is the smallest set of open transitions		satisfying the 
%rules below:
%	\end{itemize}

@@

The general case encoding relies on the causal semantics of the
algebra of BIP connectors \cite{BliSif10-causal-fmsd}.  Disregarding
the variables and data transfer, the Algebra of Connectors $\ac(P)$
\cite{BliSif07-acp-emsoft} provides a syntactic notation for the BIP
connectors.  The causal semantics of the connectors, given in terms of
the Algebra of Causal Interaction Trees $\ct(P)$, explicitates the
causal dependencies discussed above, through an encoding mapping
$\tau: \ac(P) \rightarrow \ct(P)$.  Another mapping $R: \ct(P)
\rightarrow \cru(P)$ encodes causal interaction trees into systems of
causal rules, which are Boolean implications similar to the ones in
the example above.  We briefly summarise the key definitions and
results in \app{algebras}.  The $\ac(P)$, $\ct(P)$ and $\cru(P)$
representations of the connectors in \fig{schema:ArchFailure:BIP} are
shown in \tab{connectors}.

@@

Below, we assume that, as in \fig{schema:ArchFailure:BIP}, the
interaction model is defined by a set of connectors, annotated with
Boolean guards and with update assignments.  In particular, we assume
that the guards and update assignments are well-defined for any
interaction allowed by the connector.  For example, the choice
$\NameCtrl.\PortFail\ ?\ \NameCtrl.\mathit{zone} : \topinterval$ in
the update assignement
\[
\NameTimer.\mathit{z} := \NameTimer.\mathit{z}
\cap \Big(
\bigl(
\NameCtrl.\PortFail\ ?\ \NameCtrl.\mathit{zone} : \topinterval
\bigr)
+ \NameTimer.\mathit{t} \Big)
\]
associated to the connector {\NameTimer.\PortTick \trigsynch
  (\PortFail \trigsynch \NameCtrl.\PortFail)} in
\fig{schema:ArchFailure:BIP} ensures that the assignment is
well-defined independently of whether {\NameCtrl.\PortFail}
participates or not.  Let us denote by $\gamma \subset \ac(P_A)
\times \guards{V_A} \times \assigns{V_A}$ the set of connectors in the
architecture $A$ and by $P_x$ the set of ports involved in the
connector $x \in \gamma$.  Then, we have
\begin{align*}
  \gamma &=
  \setdef{(x, g, u)}{x \in \ac(P_x), g \in \guards{V_x}, u \in
    \assigns{V_x}}
  \\
  &\simeq \setdef{(a,g,u)}{(x,g,u) \in \gamma, a \in \intsem{x}}
  = \Gamma
  \,.
\end{align*}

We first define the encoding of $\Gamma$ as $\nopri[\Gamma] \bydef{=}
\setdef{\nopri[x,g,u]}{(x,g,u) \in \gamma}$, where $\nopri[x,g,u]$ is the synchronisation vector defined by
%
\begin{equation}
  \label{eq:nopri:connector}
  \nopri[x,g,u] \bydef{=}
  \bigl(p(b_p,\export[A](p))\bigr)^{p \in P_x}
  \rightarrow
  \alpha_p
  \left[\bigwedge R(\tau(x))\bigl[\nicefrac{b_p}{p}\bigr]\right]
  \,,
\end{equation}
%
with $b_p$ a fresh Boolean variable, $\alpha_p$ a fresh name, $\tau :
\ac(P_A) \rightarrow \ct(P_A)$ and $R : \ct(P_A) \rightarrow
\cru(P_A)$ the two mappings defined in \app{transformations} and
illustrated in \tab{connectors}, $\bigl[\nicefrac{b_p}{p}\bigr]$ is
the substitution, which replaces, in the expression that preceeds it,
all occurrences of $p$ by $b_p$.

\todoLHin{The update assignment is missing from the above encoding.
  It should be $u\bigl[\nicefrac{b_p}{p}\bigr]$, but I do not see in
  \defn{pnets} where should I place it.  I think it is missing there.
  \Simon}

\begin{theorem}
  The open automaton $\semopen{\nopri[A,\partition]}$ corresponding to
  $\nopri[A,\partition]$ (see \defn{operationalSemantics}) is
  isomorphic to the LTS $\semopen{\Gamma\bigl(\cC_A, (B_D)^{D \in
      \partition}\bigr)}$ (see \defn{comp:semantics}), where, for each
  $D \in \partition$, the component $B_D \bydef{=} \tuple{\{q\}, q,
    V_D, \val{0}{D}, D, \export[D], \goesto{}}$, with
%
  \begin{align*}
    V_D &= \bigcup_{p \in D} \export[A](p)
    \,,
    &\val{0}{D}(v) &= \bot, \text{ for all $v \in V_D$}
    \,,\\
    \goesto{} &= \bsetdef{(q, p, \true, \noop, q)}{p \in D}
    \,,
    &\export[D](p) &= \export[A](p), \text{ for all $p \in D$}
    \,,
  \end{align*}
%
  and $q$ a fresh state.
\end{theorem}
%
\begin{proof}[Sketch of a proof]
  The theorem follows from the correctness of $\tau$ and $R$ shown in
  \cite{BliSif10-causal-fmsd}.
\end{proof}

\noteSBin{This theorem is very probably correct: the key question is
  whether isomorphism is, indeed, the relation to use. In absence of
  silent transitions, must be the case.}

Observe now that this encoding can be optimised to remove unnecessary
Boolean variables and loop transitions marked by \false.  For
instance, consider the ports shown in red in the second column of
\tab{connectors}.  All these ports only appear in the roots of the
corresponding causal interaction trees and, furthermore, each of these
trees only has one root (see \app{trees}).  In such cases, as in the
{\NameTimer.\PortTick \trigsynch (\PortFail \trigsynch
  \NameCtrl.\PortFail)} example discussed above, the Boolean variables
and the \false-marked loop transitions associated to these ports can
be removed.  Furthermore, the rules associated to these ports in the
corresponding systems of causal rules (see the rules in red in the
third column of \tab{connectors}) become tautological and can also be
removed from the corresponding guards in \eq{nopri:connector}.
